

# Get node trait confidence
```{bash, eval =F, engine.opts = '-l'}
conda activate nextstrain

treetime mugration --tree results/tree.nwk --states data/metadata.csv --attribute state --confidence --outdir mugration/state
treetime mugration --tree results/tree.nwk --states data/metadata.csv --attribute CO --confidence --outdir mugration/CO
treetime mugration --tree results/tree.nwk --states data/metadata.csv --attribute trap --confidence --outdir mugration/trap

```

#### * Extract Intros

```{bash, eval = F}
<!-- python3 scripts/node_data_to_table.py \ -->
<!-- --tree results/tree.nwk \ -->
<!-- --jsons results/traits.json \ -->
<!-- --include-internal-nodes \ -->
<!-- --annotations build=wnv \ -->
<!-- --output results/traits.tsv -->

```


# Header

```{r}
rm(list = ls())
pacman::p_load(treeio, ggtree, tidytree, ape, Biostrings, #phylo
               tidyverse, jsonlite, purrr)

CO_col = c("Boulder" =  "#FFF6DC",
           "Not CO" = "#96B6C5",
           "Park" = "#9E9FA5",
           "Mesa" = "#C4C1A4",
           "Weld" ="#9E9FA5", 
           "FoCo" = "#FFC6AC"
          )

zone_col = c("CO/FC/NW" = "#FFF6DC",
             "CO/FC/SE"= "#96B6C5",
             "CO/FC/NE"= "#C4C1A4",
             "CO/LV/LV"= "#9E9FA5",
             "CO/FC/SW" = "#FFC6AC",
             "CO/BE/BE" = "gray",
             "CO/FC"= "#FFC6AC",
             "CO/LV/SE"= "#9E9FA5",
             "CO/LV/NE"= "#9E9FA5")

```



#### * Import Tree and mdata

```{r}
#import tree from nextstrain
tree = read.tree("../results/tree.nwk")
mdata = read_rds("data_output/mdata_c_cmplt.rds")
fasta = readDNAStringSet("data_output/fasta_ns.fasta")
node_date0 = jsonlite::read_json("../results/branch_lengths.json")
#traits = read.delim("../results/traits.tsv")
clades = jsonlite::read_json("../results/clades.json")
mugration_state0 = read.csv("../mugration/state/confidence.csv")
mugration_CO0 = read.csv("../mugration/CO/confidence.csv")
mugration_CO_key0 = read_delim("../results/traitsCO.mugration_model.txt", delim = "\t")
mugration_trap0 = read.csv("../mugration/trap/confidence.csv")
mugration_trap_key0 = read_delim("../results/traitstrap.mugration_model.txt", delim = "\t")
```


```{r}

t2 = data.frame(f = sort(names(fasta)), m = sort(mdata$strain)) %>%
  mutate(same = f==m) %>%
  filter(same == "FALSE")

if(nrow(t2) == 0) {
  rm(t2)
}
```

# Get Clade for Lineages

```{r}
# Initialize empty lists to store node and clade membership
nodes <- character(0)
clade_membership <- character(0)

# Extract node and clade membership information
for (node_id in names(clades$nodes)) {
  node <- clades$nodes[[node_id]]
  nodes <- c(nodes, node_id)
  clade_membership <- c(clade_membership, node$clade_membership)
}

# Create a dataframe
clade_df <- data.frame(node = nodes, 
                       clade = clade_membership)

rm(nodes, node, clade_membership)

```


Get Inferred Dates for the Nodes
```{r}

node_index = which(grepl("^NODE", names(node_date0$nodes)))

node_date = node_date0$nodes[node_index]

inferred_dates = lapply(node_date, function(node) node$date)

inferred_dates = as.data.frame(inferred_dates)

inferred_dates = data.frame(t(inferred_dates))

inferred_dates$node = row.names(inferred_dates)

row.names(inferred_dates) = NULL

inferred_dates = inferred_dates %>%
  dplyr::rename("inferred_date" = `t.inferred_dates.`) %>%
  mutate(inferred_date = as.Date(inferred_date))

```

# MDATA_PHYLO:
desc: phylogenetic data (clade, trait confidence, time) & Mdata merge
```{r}
mdata_phylo = mdata %>%
  left_join(clade_df, by = c("strain" = "node")) %>%
  mutate(lineage = if_else(lineage == "unknown",
                           clade,
                           lineage))

#these should match the pruned samples
missing_mdata_phylo = mdata_phylo %>%
  filter(is.na(clade))

```

#plot of lineages
```{r}
foco = mdata_phylo 

ggplot(foco, aes(x = year, fill = lineage)) +
  geom_bar() + 
  theme_minimal()


foco = mdata_phylo %>%
  filter(CO == "FoCo")

ggplot(foco, aes(x = year, fill = lineage)) +
  geom_bar() + 
  theme_minimal()

```
# MUGRATION FUNCTION
```{r}

mugration_clean_fun = function(discreet_state, #discrete states from your mdata to help slice the mugration_key
                               mugration_key,  #mugration key output from the mugration function from timetree
                               mugration_data, #mugration matrix data from the mugration function from timetree
                               col_name){      #new colname for your discrete state of inferest
  
  n = length(unique(discreet_state)) #for slice

mugration_key = mugration_key %>%
  dplyr::slice(1:n) %>%
  separate(`Map from character to field name`, into = c("key", "inferred_state"), sep = ":") %>%
  mutate(key = row_number())


colnames(mugration_data) <- as.character(c("node",1:n))

mugration_data %>%
  pivot_longer(cols = -node, names_to = "key", values_to = "confidence") %>%
  mutate(key = as.integer(key)) %>%
  group_by(node) %>%
  filter(confidence == max(confidence)) %>%
  left_join(mugration_key, by = "key") %>%
  mutate(inferred_state = gsub("\t", "", inferred_state)) %>%
  dplyr::select(-key) %>%
    dplyr::rename(!!col_name := inferred_state)
}
```

```{r}
mugration_trap = mugration_clean_fun(discreet_state = mdata$trap,
                                     mugration_key = mugration_trap_key0,
                                     mugration_data = mugration_trap0,
                                     col_name = "inferred_trap")


mugration_CO = mugration_clean_fun(discreet_state = mdata$CO,
                                     mugration_key = mugration_CO_key0,
                                     mugration_data = mugration_CO0,
                                     col_name = "inferred_state")
mugration_CO = mugration_CO %>%
  rename(confidence_state = confidence)

mugration_trap = mugration_trap %>%
  rename(confidence_trap = confidence)

```


#### If missing lineages then extract lineages
```{r}
miss_lin = mdata %>% filter(lineage == "unknown") %>% nrow()

if(miss_lin > 0) {
  #figure out how to add lineage
}

```


# convert tree to df add mdata convert back to tree
```{r}

df_tree = tidytree::as_tibble(tree)

df_tree2 = left_join(df_tree, mugration_CO, by = c("label" = "node")) %>%
           left_join(mugration_trap, by = c("label" = "node")) %>%
           left_join(inferred_dates, by = c("label" = "node")) %>%
           left_join(clade_df, by = c("label" = "node")) %>%
           left_join(mdata, by = c("label" = "strain")) %>%
           mutate(tip = !str_detect(label, "^NODE")) %>%
           mutate(inferred_date = dplyr::if_else(is.na(inferred_date),
                                         date,
                                         inferred_date)
                  ) 


tree2 = as.treedata(df_tree2)

p_tree = ggtree(tree2) +
  geom_nodepoint(aes(color = inferred_state)) +
 # geom_text(aes(label=node), ,size = 2, hjust=-.3) +
  #scale_color_manual(values = CO_col) +
  # geom_tippoint(aes(color = CO)) +
  # scale_color_manual(values = CO_col) +
  geom_treescale()

p_tree
```


#starting from node method 2
```{r}
tree_data = df_tree2
trait = "FoCo"

#get the node numbers for your inferred_state trait of interest
intro_node = tree_data %>% 
  filter(tip == F) %>%
  filter(inferred_state == trait)

#extract the node numbers from the dataframe
intro_node_num = intro_node$node

#extract the inferred date to filter to the intro node from the tip 
intro_node_date = intro_node$inferred_date

#get list of nodes of parents a of your node of inferred state of interest
intro_node_list <- map(intro_node_num, ~parent(tree_data, .x))

#add parent nodes inferred state and inferred date to a dataframe for filtering
parent_node = lapply(intro_node_list, function(df) df[, c("node", "inferred_state", "inferred_date")])

#remove nodes where parent interred state is the same as node of interest. AKA keep nodes with a change in the inferred_state aka introduction
parent_node = do.call(rbind, parent_node) %>% 
  dplyr::rename("parent_node" = node,
                "parent_inferred_state" = inferred_state,
                "parent_inferred_date" = inferred_date) %>%
  mutate(intro_node_num = intro_node_num,
         intro_node_date = intro_node_date) %>%
  filter(parent_inferred_state != trait)

#get numbers in a list for the for loop
node_num = parent_node$intro_node_num

#get all the j tip sequences that are in the intro nodes i filter for above
intro_df_list = map(node_num, ~offspring(tree_data, .x, tiponly = T))
                    

#get add the intro node i of the j tips
intro_df_list2 = list()

for(i in 1:(length(intro_df_list))){
  intro_df_list2[[i]]= cbind(intro_df_list[[i]], parent_node[i, c("intro_node_num", "intro_node_date")])
}


#intro_df_list2 = sample(intro_df_list2, 3)

```

FOR LOOP TO FILTER TIPS WITH K ANCESTORS NOT EQUAL TO THE INFERRED STATE
```{r}
tree_data = df_tree2
trait = "FoCo"

intro_df_list3 = list()

for(i in seq_along(intro_df_list2)){
  
#get node date for the intro node i  
node_date = unique(intro_df_list2[[i]]$intro_node_date)
  
#get the number of tip j node for offspring of node i

#j = 1

tip_ij = as.numeric(intro_df_list2[[i]][,"node"])

  tip_ij_anc=  tip_ij %>%
    map(~ancestor(tree_data, .x))  %>% 
    map(~filter(.x, inferred_date >= node_date))

#add the name of tip_ij to match it with its ancestor nodes 
for(j in 1:length(tip_ij)) {

  tip_ij_anc[[j]] = mutate(tip_ij_anc[[j]], tip_ij = tip_ij[j])
}

#function for true false do all of the inferred state equal the inferred state of interest?
all_ij_trait_ancestor = function(df, x){
  all(grepl(trait, df$inferred_state))
}

#only keep the tips with ancestor nodes that are all the inferred state of interest indicating it is a continued transmission and didn't leave that state
tip_ij_anc2 = keep(tip_ij_anc, all_ij_trait_ancestor)

#extract the tip_ij whos ancestor k were all trait of interest (FoCo)
tip_ij_keep = map_dbl(tip_ij_anc2, ~(unique(.x$tip_ij)))

intro_df_list3[[i]] = intro_df_list2[[i]] %>%
  mutate(tip_keep = if_else(node %in% tip_ij_keep,T,F))

}

intro_GT2 = map_df(intro_df_list3, ~cbind(.x)) 
                    
map_dbl(intro_df_list3, ~filter(.x, tip_keep == T) %>% nrow()) 
```

Get single tips of trait intro 
```{r}
#get the node numbers for your inferred_state trait of interest
intro_node = tree_data %>% 
  filter(tip == T) %>% #changed from above
  filter(inferred_state == trait)

#extract the node numbers from the dataframe
intro_node_num = intro_node$node

#extract the inferred date to filter to the intro node from the tip 
intro_node_date = intro_node$inferred_date

#get list of nodes of parents a of your node of inferred state of interest
intro_node_list <- map(intro_node_num, ~parent(tree_data, .x))

#add parent nodes inferred state and inferred date to a dataframe for filtering
parent_node = lapply(intro_node_list, function(df) df[, c("node", "inferred_state", "inferred_date")])

#remove nodes where parent interred state is the same as node of interest. AKA keep nodes with a change in the inferred_state aka introduction
parent_node = do.call(rbind, parent_node) %>% 
  dplyr::rename("parent_node" = node,
                "parent_inferred_state" = inferred_state,
                "parent_inferred_date" = inferred_date) %>%
  mutate(intro_node_num = intro_node_num,
         intro_node_date = intro_node_date) %>%
  filter(parent_inferred_state != trait)


node_num = parent_node$intro_node_num

single_intro = tree_data %>% 
  filter(node %in% node_num) %>%
  mutate(intro_node_num = node,
         intro_node_date = date,
         tip_keep = T)
```


# COMBINE INTROS ADD DATA TO FULL TREE
```{r}
intro_samples = rbind(single_intro, intro_GT2) %>%
  filter(tip_keep == T) 

t = intro_samples %>%
  dplyr::select(node, intro_node_num, intro_node_date, tip_keep)

df_tree3 = df_tree2 %>%
  left_join(t, by = "node") %>%
  mutate(intro_tip_state = if_else(!is.na(tip_keep),
                             state,
                             NA)
         )
```

# Write files
```{r}
saveRDS(intro_samples, "data_output/intro_samples.rds")
saveRDS(df_tree3, "data_output/df_tree2.rds")
```



